taxize - taxonomic search and retrieval in R

### 2013-05-16 inst/doc/taxize_withcode.Rnw

There are a large suite of applications developed around the problem of searching for, resolving, and getting higher taxonomy for species names. For example, \bf{Linnaeus} \url{http://linnaeus.sourceforge.net/} provides the ability to search for taxonomic names in documents and normalize those names found. In addition, there are many web interfaces to search for and normalize names such as Encyclopedia of Life's Global Names Resolver \url{http://resolver.globalnames.org/}, uBio tools \url{http://www.ubio.org/index.php?pagename=sample_tools}, and iPlant's Taxonomic Name Resolution Service \url{http://tnrs.iplantcollaborative.org/}.

@@

Some of the data sources taxize interacts with require authentication. That is, in addition to the search terms the user provides (e.g., \emph{Homo sapiens}), the data provider requires an alphanumeric identification key so that they can better manage their servers, collect analytics, and shut down users that abuse the API. The services that do require an API key in taxize are: Encyclopedia of Life (EOL), the Universal Biological Indexer and Organizer (uBio), Tropicos, and Plantminer. You can easily obtain an API key by visiting the website of each service (see (Table~\ref{tab:a}) for links to each site). There are two ways of using your API keys. First, you can pass in your API key in a function call (e.g., \emph{ubio_namebank(srchName='Ursus americanus', key='your_alphanumeric_key')}). Second, you can store your API keys in your .Rprofile file. On a Mac this file is at \emph{/yourhomefolder/.Rprofile}; on a Windows machine at \emph{/yourhomefolder/.Rprofile}; and on Linux at \emph{/yourhomefolder/.Rprofile}. This is an hidden file, so open up this file in your terminal (e.g., \emph{open .Rprofile}), and add the API key as an entry like \emph{options(myapikey = 'your_alphanumeric_key')}. We recommend the second option as it simplifies function calls.

@@

\begin{table}[!ht]
\caption{Data sources used in taxize, tasks available, and links to them}\label{tab:a} % title of Table
\begin{tabular}{|l|cccc|l|}
\hline
Source name & Name search & Name resolution & Phylogeny & Sequences & URL  \\
\hline
Encyclopedia of Life & Yes & See GNR below & No & No & \url{http://eol.org/} \\
Integrated Taxonomic Information System & Yes & Synonyms & No & No & \url{http://www.itis.gov/} \\
iPlant Taxonomic Name Resolution Service & Yes & Yes & No & No & \url{http://bit.ly/16dHkBy} \\
Phylomatic & No & No & Yes & No & \url{ http://bit.ly/P0pjMz} \\
uBio & Yes & Yes & No & No & \url{http://www.ubio.org/} \\
Global Names Resolver & Yes & Yes & No & No & \url{http://bit.ly/11R3Pbr} \\
Global Names Index & Yes & No & No & No & \url{http://bit.ly/11R3RQB} \\
IUCN Red List & Yes & No & No & No & \url{http://bit.ly/11R3RQC} \\
Tropicos & Yes & Yes & No & No & \url{http://www.tropicos.org/} \\
Plantminer & Yes & No & No & No & \url{http://www.plantminer.com/} \\
The Plant List & Yes & Yes & No & No & \url{http://www.theplantlist.org/} \\
Catalogue of Life & Yes & Yes & No & No & \url{http://bit.ly/11R3S75} \\
National Center for Biotechnology Information & Yes & X & Yes \footnote{Web only, no API, see \url{http://1.usa.gov/11R446a}} & Yes & \url{http://www.ncbi.nlm.nih.gov/} \\
\hline
\end{tabular}
\end{table}

@@

<<resolvenames, eval=TRUE>>=
# Here, we are searching for two misspelled names
temp <- gnr_resolve(names = c("Helianthos annus", "Homo saapiens"), returndf = TRUE)

@@

<<tnrs, eval=TRUE>>=
# A list of species names, some of which are misspelled
mynames <- c("Helianthus annuus", "Pinus contort", "Poa anua", "Abis magnifica",
		"Rosa california", "Festuca arundinace", "Sorbus occidentalos","Madia sateva")

@@

<<itisclass, eval=TRUE>>=
specieslist <- c("Abies procera","Pinus contorta")
classification(get_tsn(specieslist, "sciname"))
@

@@

<<taxname, eval=TRUE>>=
# Using ITIS 
tax_name(query="Helianthus annuus", get="family", db="itis")

@@

In the previous section we used the function \emph{get_tsn} to get a classification for two plant species. Below are a few examples. When you run these examples in R, you are presented with a command prompt asking for the row that contains the name you would like back; that output is not printed below fro brevity.

<<interactive, eval=FALSE, echo=TRUE>>=
# In this example, this search term has many matches. The function returns a data.frame of the matches, and asks for user input. Enter the row number of the name you want. 
get_tsn(searchterm="Elisabethae", searchtype="sciname")

@@

<<phylomaticphylogeny, eval=TRUE, dev='png', out.width='4in', out.height='4in', dpi=100>>=
# Input the taxonomic names
taxa <- c("Poa annua", "Abies procera", "Helianthus annuus")

@@

Behind the scenes the function \emph{phylomatic\_tree} retrieves a Taxonomic Serial Number (TSN) from ITIS for each species name, then a string is created for each species like this \emph{poaceae/oryza/oryza_sativa} (with format 'family/genus/genus_epithet'). These strings are submitted to the Phylomatic API, and if no errors occur, a phylogeny in newick format is returned. The \emph{phylomatic\_tree()} function also cleans up the newick string and converts it to an \textbf{ape} \emph{phylo} object. The output from \emph{phylomatic\_tree()} is a \emph{phylo} object, which can be used for plotting and phylogenetic analyses.

@@

<<downstream, eval=TRUE>>=
# Search for all species downstream from Apis
col_downstream(name="Apis", downto="Species")[[1]]
@

@@

<<iucn, eval=TRUE>>=
ia <- iucn_summary(c("Panthera uncia", "Lynx lynx"))
iucn_status(ia)
@

@@

<<genbank, eval=TRUE>>=
# Serch for available genes in GenBank.
out <- get_genes_avail(taxon_name="Umbra limi", seqrange = "1:2000", getrelated=FALSE)

# Get a list of available genes, removing non-unique records.
unique(out$genesavail)

>>>>

There are a large suite of applications developed around the problem of searching for, resolving, and getting higher taxonomy for species names. For example, \textbf{Linnaeus} \url{http://linnaeus.sourceforge.net/} provides the ability to search for taxonomic names in documents and normalize those names found. In addition, there are many web interfaces to search for and normalize names such as Encyclopedia of Life's Global Names Resolver \url{http://resolver.globalnames.org/}, uBio tools \url{http://www.ubio.org/index.php?pagename=sample_tools}, and iPlant's Taxonomic Name Resolution Service \url{http://tnrs.iplantcollaborative.org/}.

@@

Some of the data sources taxize interacts with require authentication. That is, in addition to the search terms the user provides (e.g., \emph{Homo sapiens}), the data provider requires an alphanumeric identification key so that they can better manage their servers, collect analytics, and shut down users that abuse the API. The services that do require an API key in taxize are: Encyclopedia of Life (EOL), the Universal Biological Indexer and Organizer (uBio), Tropicos, and Plantminer. You can easily obtain an API key by visiting the website of each service (see (Table~\ref{tab:a}) for links to each site). There are two ways of using your API keys. First, you can pass in your API key in a function call (e.g., \emph{ubio\_namebank(srchName='Ursus americanus', key='your\_alphanumeric\_key')}). Second, you can store your API keys in your .Rprofile file. On a Mac this file is at \emph{/yourhomefolder/.Rprofile}; on a Windows machine at \emph{/yourhomefolder/.Rprofile}; and on Linux at \emph{/yourhomefolder/.Rprofile}. This is an hidden file, so open up this file in your terminal (e.g., \emph{open .Rprofile}), and add the API key as an entry like \emph{options(myapikey = 'your\_alphanumeric\_key')}. We recommend the second option as it simplifies function calls.

@@

\begin{table}[!ht]
\caption{Data sources used in taxize, tasks available, and links to them}
\label{tab:a} % title of Table
\begin{tabular}{|l|cccc|l|}
\hline
Source name & Name search & Name resolution & Phylogeny & Sequences & URL  \\
\hline
Encyclopedia of Life & Yes & See GNR below & No & No & \url{http://eol.org/} \\
Integrated Taxonomic Information System & Yes & Synonyms & No & No & \url{http://www.itis.gov/} \\
iPlant Taxonomic Name Resolution Service & Yes & Yes & No & No & \url{http://bit.ly/16dHkBy} \\
Phylomatic & No & No & Yes & No & \url{ http://bit.ly/P0pjMz} \\
uBio & Yes & Yes & No & No & \url{http://www.ubio.org/} \\
Global Names Resolver & Yes & Yes & No & No & \url{http://bit.ly/11R3Pbr} \\
Global Names Index & Yes & No & No & No & \url{http://bit.ly/11R3RQB} \\
IUCN Red List & Yes & No & No & No & \url{http://bit.ly/11R3RQC} \\
Tropicos & Yes & Yes & No & No & \url{http://www.tropicos.org/} \\
Plantminer & Yes & No & No & No & \url{http://www.plantminer.com/} \\
The Plant List & Yes & Yes & No & No & \url{http://www.theplantlist.org/} \\
Catalogue of Life & Yes & Yes & No & No & \url{http://bit.ly/11R3S75} \\
National Center for Biotechnology Information & Yes & X & Yes \footnote{Web only, no API, see \url{http://1.usa.gov/11R446a}} & Yes & \url{http://www.ncbi.nlm.nih.gov/} \\
\hline
\end{tabular}
\end{table}

@@

<<resolvenames, eval=TRUE, cache=TRUE>>=
# Here, we are searching for two misspelled names
temp <- gnr_resolve(names = c("Helianthos annus", "Homo saapiens"), returndf = TRUE)

@@

<<tnrs, eval=TRUE,cache=TRUE>>=
# A list of species names, some of which are misspelled
mynames <- c("Helianthus annuus", "Pinus contort", "Poa anua", "Abis magnifica",
		"Rosa california", "Festuca arundinace", "Sorbus occidentalos","Madia sateva")

@@

<<itisclass, eval=TRUE,cache=TRUE>>=
specieslist <- c("Abies procera","Pinus contorta")
classification(get_tsn(specieslist, "sciname"))
@

@@

<<taxname, eval=TRUE,cache=TRUE>>=
# Using ITIS 
tax_name(query="Helianthus annuus", get="family", db="itis")

@@

In the previous section we used the function \emph{get\_tsn} to get a classification for two plant species. Below are a few examples. When you run these examples in R, you are presented with a command prompt asking for the row that contains the name you would like back; that output is not printed below fro brevity.

<<interactive, eval=FALSE, echo=TRUE, cache=TRUE>>=
# In this example, this search term has many matches. The function returns a data.frame of the matches, and asks for user input. Enter the row number of the name you want. 
get_tsn(searchterm="Elisabethae", searchtype="sciname")

@@

<<phylomaticphylogeny, eval=TRUE, dev='pdf', out.width='4in', out.height='4in', cache=TRUE>>=
# Input the taxonomic names
taxa <- c("Poa annua", "Abies procera", "Helianthus annuus")

@@

Behind the scenes the function \emph{phylomatic\_tree} retrieves a Taxonomic Serial Number (TSN) from ITIS for each species name, then a string is created for each species like this \emph{poaceae/oryza/oryza\_sativa} (with format 'family/genus/genus\_epithet'). These strings are submitted to the Phylomatic API, and if no errors occur, a phylogeny in newick format is returned. The \emph{phylomatic\_tree()} function also cleans up the newick string and converts it to an \textbf{ape} \emph{phylo} object. The output from \emph{phylomatic\_tree()} is a \emph{phylo} object, which can be used for plotting and phylogenetic analyses.

@@

<<downstream, eval=TRUE, cache=TRUE>>=
# Search for all species downstream from Apis
col_downstream(name="Apis", downto="Species")[[1]]
@

@@

<<iucn, eval=TRUE, cache=TRUE>>=
ia <- iucn_summary(c("Panthera uncia", "Lynx lynx"))
iucn_status(ia)
@

@@

<<genbank, eval=TRUE, cache=TRUE>>=
# Serch for available genes in GenBank.
out <- get_genes_avail(taxon_name="Umbra limi", seqrange = "1:2000", getrelated=FALSE)

# Get a list of available genes, removing non-unique records.
# unique(out$genesavail)
